Performance for the masses: Experiments with A Web Based Architecture to Harness Volunteer Resources for Low Cost Distributed Evolutionary Computation

### 2016-02-01 GECCO/volunteer-clouds-gecco16.tex
Reduced it to 8 pages
Please @unintendedbear @mariosky @fergunet check it for consistency, explained things (or not) and anything you can see. GECCO is a very difficult venue, we really have to nail it.

Using the browser as a computing resource in volunteer computing presents several advantages, 
but it remains a challenge to harness the browser's capabilities and
to model the user's behavior so that those capabilities can be
leveraged optimally. In this paper, we present the results of several
experiments with a volunteer computing framework called {\sf NodIO}
that is easily deployable in using free cloud resources. We use 
%Pablo: "deployable in using free cloud resources" is correct?
different browser configurations to find out which one is 
able to get the user to lend more computing cycles and test different
problems to check the influence it has on said performance.
% Pablo: so the objective is "to determine which browser conf. is able
% to get more computing power from the users", it isn't? 
% (to keep this in mind the rest of the paper). What is a "browser configuration"?
% (explain it with a phrase in the abstract)
From these experiments we can already draw some conclusions, besides
the fact that volunteer computing can be a valuable computing
resource: the user has to be kept engaged in some way to be able to
use as many computing cycles as possible, the client has to be built
to use the computer capabilities fully, and there is some common model
to all sets of users, following a common statistical distribution. As
a pragmatic result, establishing trust with the user by open sourcing
not only the software, but also the data, is essential to have a user %Pablo: remember, is "yiuser" :P
base as wide as possible.

@@

%PABLO: He añadido el siguiente párrafo para el issue #31
There exist other volunteer computing frameworks written in compiled languages,
such as C++ or Java, that, in principle, can achieve better performance than
JavaScript \cite{2015arXiv151101088M}. However, one of the advantages of our architecture is that
clients do not need to install anything in their machines, they just
need to open a web page  
% Somewhat redundant 
% Javascript is the only language that can be executed in this way without any installation, 
% also removed the default, redundant with next sentence
in a compatible browser, such as Chrome, Safari or Firefox, which are
available in almost every desktop or mobile device. Other advantage is
that the source code can be directly examined by volunteers, because it is not
compiled as other options such as Java Applets, ActionScript or
ActiveX.
Also, as we will discuss in this paper, it can provide some extra
features such as enhanced security or operating system independence.

@@

The main problem with using Java and all the other browser-embedded
technologies is that they are not {\em universal} in the
sense that an extra component, namely, the Java virtual machine or another
virtual environment has to be installed in the browser.
Also these virtual environments have become targets of malicious attacks; 
for instance numerous vulnerabilities have been discovered in 
the proprietary Adobe Flash runtime environment and
video player \cite{ford2009analyzing,watanabe2010new}. The need for
these additional environments has been mitigated with the
adoption of the HTML5 standard \cite{anthes2012html5}.
JavaScript however, has become a set of standards
\cite{ECMA-262} that include the language and its components.

JavaScript can be used either for unwitting, that is, stealth,
\cite{unwitting-ec} or volunteer 
\cite{langdon:2005:metas,gecco07:workshop:dcor} distributed
evolutionary computation and it has been used ever since by several
authors, including more recent efforts
\cite{duda2013distributed,DBLP:journals/corr/abs-0801-1210} 
that even
used the client's GPU \cite{duda2013gpu}. Even if, as said above, Java
is not the best option, many other researchers have
used Java \cite{chong:1999:jDGPi} and others have gone away from the
server-based paradigm to embrace peer to peer systems
\cite{jin2006constructing,10.1109/ICICSE.2008.99,DBLP:conf/3pgcic/GuervosMFEL12}. These
computing 
platforms avoid single points of failure (the server) but, since they
need a certain amount of infrastructure installed to start, the
threshold to join them is much higher. % (Paloma) Do yo
                                % mean... higher? A low threshold
                                % means that is easy to happen
                                % something.
%Correct. But maybe in this new version we should eliminate this
%paragraph altogether. 
% The next paragraph has been suppresed.

JavaScript is also compatible with the functional programming paradigm \cite{Cousineau1998,MacLennan1990,Thompson1996}.
According to \cite{swanresearch2015}, the benefit of using this kind of paradigm in
metaheuristics optimization is that it allows an improvement in communicability,
reproductibility, interoperability, automated assembly, knowledge engineering,
and efficiency. In the work presented in this paper, JavaScript has been used throughout the full
stack, which has the advantage of using the self-same lines of code for
writing an evolutionary algorithm for the
desktop or command line and browser clients, both based in the EC
library {\sf NodEO} \cite{DBLP:conf/gecco/GuervosVGES14}. This saves time 
when writing code and leverages 
the expertise acquired
% (Paloma) Not sure if "acquired" should be written before
% "expertise"... sounds better to me, anyways
% Don't know what you mean. Write it down and we'll see - JJ
 in the library itself.

@@

As far as we know, this paper presents one of the few experimental
results using a volunteer computing system that uses  % (Paloma) Change to "which" in order to not repeating "that" all the time
computational resources that are as dissimilar as smartphones and
powerful laptops 
or workstation computers in a research center. This will be described next.

@@

Initial results using the same code base and a single computer have
been described in \cite{2016arXiv160101607Manom}. They prove that a bigger population
contributes to diversity and thus speeds up the solution
\cite{DBLP:conf/lion/LaredoDFGB13}. The volunteer computing 
experiments that we will describe next do not and can not have the
same conditions, but
the baseline is that if they eventually take longer than a basic
desktop, their interest will be purely academic, although they can
complement them. We will try and
prove next that that target performance can be achieved by carrying
out several experiments in different conditions. Second, they also act
as a basic performance test for the server, and also prove that the
basic server can be used for many kinds of different setups, as long
as they follow the REST protocol to deposit and obtain individuals
from the pool. It is obvious that, in this case, the server is not
really needed, since there is a single client and no interaction
between them, but there is a cost in sending and obtaining chromosomes
from the server so it is included for the sake of fairer comparison.

Initial experiments were set up using the OpenShift
PaaS, which provides a free tier within which these runs could be performed. That way, and as required, the hardware
and cloud cost for this experiment were zero. Experiments were
announced through a post in Twitter and other social networks, and
results were published here \cite{DBLP:conf/gecco/GuervosG15anom}. For the
purpose of this paper, we repeated the announcement several times
through the month of April and then by the beginning of August. All
in all, we have the set of runs with the characteristics shown in
Table \ref{tab:summary:os}. In general, every experiment took several
days. No particular care was taken about the time of the announcement
or the particular wording. Every {\em experiment} consisted in running
until the solution of the 40-trap problem was found. When the correct
solution was sent to the server, the counter was updated and the pool
of solutions reset to the void set. There was no special intention to wait
until all clients had finished, thus it might happen that, in fact,
the islands running in the browser {\em spill} from one experiment to
the next. However, previous experiments have proven that the influence
of these islands in the next experiment is indeed negligible.
%
\begin{table*}
\caption{Summary of time per run, number of IPs and number of PUTs per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccccc}
\hline
Date & Experiments &\#IPs &\#IPs &  \#IPs & Time (s) &  \#PUTs & $<$ 69s & $<$ 3.46s & Inter-experiment\\
     & \# & Different & Median & Max & Median &  Median & & & correlation \\ 
\hline
4/4 & 57 & 191 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 & 231 & 559 & 5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 97 & 179 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%
The table shows that every run included more than 50 experiments. The
number of different IPs intervening in them varied from more than one
hundred to more than five hundred in the second experiment.

A summary of the results of each run is also shown in Table
\ref{tab:summary:os}, which shows the median number of IPs
intervening in each experiment,  median time needed
to finish the experiment, median number of HTTP PUTs per IP, and
then two columns showing the percentage of experiments that took less
than the two baseline experiments shown in the introduction to this
section. The first striking result is that in all cases, 50\% of the
experiments involved 5 or less IPs. This is consistent with results
previously obtained \cite{DBLP:conf/gecco/GuervosG15anom} which found 6 to be
the usual number of IPs that participated in an experiment. The
maximum number of different IPs for each experiment is also in the
same range and of the order of 10, which is also consistent with
prior work. The median time has a big range of variation, but 50\% of
the time takes less than several minutes, from around 4 minutes in the
best case to roughly 2/3 of an hour in the worst case. That figure is
not competitive, {\em a priori}, with the baseline experiment, that is
why it is interesting to look at the columns which measure
the percentage of experiments in which the solution was found in less
time than the baseline. It is similar in all three cases, with around
20\% on average for the first baseline experiment, roughly 3\% in the
second case. The comparison is not totally fair: this experiment takes
place in a browser, and in fact most of the time is devoted to
presenting the charts; the server is also not the same since it is run in a
PaaS instead of a high-performance desktop computer, and it is also a
remote server, not a local server as in the baseline experiment. But
even in this case, we were looking for creating a framework that
allowed us to run evolutionary computation experiments {\em faster} than
in our very own computer. And the results indicate that although it will
happen in 1 out of roughly 5 or roughly 20 cases, depending on what baseline
experiment you choose, it will not happen always or even most of
the time. One of the problems is shown in the last column, which
indicates the statistical correlation between the number of IPs
participating in one experiment and the next. In all cases,
correlation is so low as to affirm that there is almost no
relationship between them, that is, computing nodes are not {\em
  staying} after the solution has been found. This is a feature of
this implementation: the user has to actively reload the page to make
the evolutionary algorithm start again. Thus, the user can decide if to 
continue sharing resources or not. % (Paloma) Feature or disadvantage?
However, it puts a probably            % (Mario) Feature if you only want to share your CPU a little  
unnecessary upper limit on the time, or number of experiments, that a user
will voluntarily perform.

\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
\caption{Duration of experiments vs. number of different IPs (nodes)
  participating in it, with averages and standard deviation shown as
  black dots and lines; in the case there is a single dot, there was a single
  experiment in which so many computers participated (for instance, 16
  computers in the experiment in the far left or 29 in the middle
  one). 
Shade of gray or color indicate how many experiments included that many unique IPs,
so lighter (bluer) shade for a column of dots indicates that a particular number
of computers happened less frequently, while darker shade or closer to
red means more frequency. 
From left to right, experiments 4/4, 4/24 and 7/31.}
\label{fig:duration}
\end{figure}
% ../data/time-vs-IPs-openshift.R
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color (or gray value) is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
% computed via ../data/time-vs-IPs-openshift.R
\end{figure}
%
We will have to analyze experimental data a bit further to find out why
this happens and also if there are some patterns in the three sets of
experiments. An interesting question to ask, for instance, is if
by adding more computers makes the experiment take less. In fact, as
shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
not seem to contribute to decreasing the time needed to finish the
experiment. However, the cause-effect relationship is not clear at
all. It might be the opposite: since experiments take longer to finish
and might in fact be abandoned with no one contributing for some time,
that increases the probability of someone new joining them. In fact,
with experiments taking a few seconds and due to the way the
experiments are announced, it is quite difficult that several
volunteers join in in such a short period of time, even more if we take
into account that volunteers are not {\em carried over} from previous
experiments. This implies that it would be convenient to use a problem
of a bigger size to check this hypothesis as we have done in the next
experiment. 
%  however, at this point we
% have not found this convenient since there are several other issues
% that have to be solved, as it will be shown next. %Creo que este párrafo
%saldría sobrando ahora que hicmios un experimento más grande

@@

Although we think that, in general, the results presented in this
paper are independent of the problem chosen, it is true that even if
it is a difficult problem for evolutionary algorithms, it takes a few
seconds with the right settings to solve. This makes it difficult also
for this kind of asynchronous framework, but it would be interesting
to check the performance for other computationally demanding
tasks, especially ones that cover the middle ground between a problem
finishing in seconds or minutes (Trap) and weeks (Rastrigin F15). 
Besides, the two versions of the framework include as an
algorithmic variant using random population size. We do not think this
has had a big influence on the results and in fact this is not noticed
for F15, which is the second version tested, but it would be interesting to
measure exactly what this influence has been. In general there are
many issues with the evolutionary algorithm implementation itself,
including using different, or adaptive, policies for inserting and
sending individuals to the pool as was done in \cite{araujo2008mam},
using different policies for population initialization, and also the
incorporation of high-speed local resources to the pool to check what
would be the real influence of the volunteer pool to the final
performance.

>>>>

Using the browser as a computing resource in volunteer computing presents several advantages, 
but it remains a challenge to harness the browser's capabilities and
to model the user's behavior so that those capabilities can be
leveraged optimally. In this paper, we present the results of several
experiments with a volunteer computing framework called {\sf NodIO}
that is easily deployable in using free cloud resources. We use 
%Pablo: "deployable in using free cloud resources" is correct?
% why not? - JJ
different implementations to find out which one is 
able to get the user to lend more computing cycles and test different
problems to check the influence it has on said performance.
% Pablo: so the objective is "to determine which browser conf. is able
% to get more computing power from the users", it isn't? 
% (to keep this in mind the rest of the paper). What is a "browser configuration"?
% (explain it with a phrase in the abstract)
% No, we don't. Please read the paper!!!! - JJ
From these experiments we can already draw some conclusions, besides
the fact that volunteer computing can be a valuable computing
resource: the user has to be kept engaged in some way to be able to
use as many computing cycles as possible, the client has to be built
to use the computer capabilities fully, and, finally, the user
contributions follow a common statistical distribution. It is also
essential to be as open as possible with software and data.

@@

%PABLO: He añadido el siguiente párrafo para el issue #31
There exist other volunteer computing frameworks written in compiled languages,
such as C++ or Java, that, in principle, can achieve better performance than
JavaScript. However, one of the advantages of our architecture is that
clients do not need to install anything in their machines, they just
need to open a web page  
% Somewhat redundant 
% Javascript is the only language that can be executed in this way without any installation, 
% also removed the default, redundant with next sentence
in a compatible browser, such as Chrome, Safari or Firefox, which are
available in almost every desktop or mobile device. Other advantage is
that the source code can be directly examined by volunteers, because it is not
compiled as other options such as Java Applets, ActionScript or
ActiveX.
Also, as we will discuss in this paper, it can provide some extra
features such as enhanced security or operating system independence.

@@

The main problem with using Java and all the other browser-embedded
technologies is that they are not {\em universal} in the
sense that an extra component, namely, the Java virtual machine or another
virtual environment has to be installed in the browser.
Also these virtual environments have become targets of malicious attacks; 
for instance numerous vulnerabilities have been discovered in 
the proprietary Adobe Flash runtime environment and
video player \cite{ford2009analyzing,watanabe2010new}. The need for
these additional environments has been mitigated with the
adoption of the HTML5 standard \cite{anthes2012html5}.
JavaScript however, has become a set of standards that include the language and its components.

JavaScript can be used either for unwitting, that is, stealth,
\cite{unwitting-ec} or volunteer 
\cite{langdon:2005:metas,gecco07:workshop:dcor} distributed
evolutionary computation and it has been used ever since by several
authors, including more recent efforts
\cite{duda2013distributed,DBLP:journals/corr/abs-0801-1210} 
that even
used the client's GPU \cite{duda2013gpu}. 
%Also a bit out of scope. Have to reduce the paper. 
% Even if, as said above, Java
% is not the best option, many other researchers have
% used Java \cite{chong:1999:jDGPi} and others have gone away from the
% server-based paradigm to embrace peer to peer systems
% \cite{jin2006constructing,10.1109/ICICSE.2008.99,DBLP:conf/3pgcic/GuervosMFEL12}. These
% computing 
% platforms avoid single points of failure (the server) but, since they
% need a certain amount of infrastructure installed to start, the
% threshold to join them is much higher. % (Paloma) Do yo
                                % mean... higher? A low threshold
                                % means that is easy to happen
                                % something.
%Correct. But maybe in this new version we should eliminate this
%paragraph altogether. 
% The next paragraph has been suppresed.

%Also suppressing this one which is probably not needed in this paper
%- JJ
% JavaScript is also compatible with the functional programming paradigm \cite{Cousineau1998,MacLennan1990,Thompson1996}.
% According to \cite{swanresearch2015}, the benefit of using this kind of paradigm in
% metaheuristics optimization is that it allows an improvement in communicability,
% reproductibility, interoperability, automated assembly, knowledge engineering,
% and efficiency. In the work presented in this paper, JavaScript has been used throughout the full
% stack, which has the advantage of using the self-same lines of code for
% writing an evolutionary algorithm for the
% desktop or command line and browser clients, both based in the EC
% library {\sf NodEO} \cite{DBLP:conf/gecco/GuervosVGES14}. This saves time 
% when writing code and leverages 
% the expertise acquired
% % (Paloma) Not sure if "acquired" should be written before
% % "expertise"... sounds better to me, anyways
% % Don't know what you mean. Write it down and we'll see - JJ
%  in the library itself.

@@

As far as we know, this paper presents one of the few experimental
results using a volunteer computing system which uses  % (Paloma) Change to "which" in order to not repeating "that" all the time
very diverse computational resources; these will be described next.

@@

We set up different experiments by deploying the server in a cloud
service, which changed in every one. This server runs a web service
which stores chromosomes sent from clients and sends them back. It is
an open source system that has been described extensively in
\cite{2016arXiv160101607Manom}. The client runs an evolutionary
algorithm written in JavaScript, that after a number of generations
sends back the best individual to the server, requesting at the same
time a random individual of those stored in the server. The experiment
starts by setting up a website that is sent by the server and announcing it on several social
networks so that people contribute to them by going to the page and
staying there for a while.

\subsection{Initial tests}

Initial tests with an evolutionary algorithm solving the 40-Trap
problem were set up using the free PaaS (Platform as a
Service) OpenShift; they are described in
Table \ref{tab:summary:os}. In general, every tests took several
days and during it the problem was solved more than 50 times with
several hundreds different IPs participating in it; the
evolutionary algorithm stopped when the solution was found. 
%
\begin{table*}
\caption{Summary of time per run, number of IPs and number of PUTs per IP in the initial runs. \label{tab:summary:os}}
\begin{center}
\begin{tabular}{l|ccccccccc}
\hline
Date & Experiments &\#IPs &\#IPs &  \#IPs & Time (s) &  \#PUTs & $<$ 69s & $<$ 3.46s & Inter-experiment\\
     & \# & Different & Median & Max & Median &  Median & & & correlation \\ 
\hline
4/4 & 57 & 191 & 5 & 16 & 2040 & 18 & 14.29\% & 5.36\% & 0.0082 \\
4/24 & 231 & 559 & 5 & 29 & 732 & 11 & 47.39\% & 3.91\% & 0.0934\\
7/31 & 97 & 179 & 5 & 14 & 260 & 23 & 27.08\% & 1.04\%  & 0.1741\\
\hline
\end{tabular}
\end{center}
\end{table*}
%

The last column shows the percentage of experiments in which running the
experiment in the browser takes {\em less} time than in the
desktop. Although volunteer computing systems do not exclude desktop
computing and, in fact, the server can handle any client, not only
those running JavaScript on the browser, we include it here for the
sake of comparison and to show that, being a free resource, it can
anyway make a valid and valuable contribution to any distributed or
desktop computing experiment. An interesting result is also that the
correlation between users of an experiment and the next is
non-existent, that is, computing nodes are not {\em
  staying} after the solution has been found. This is an issue that we
will address in this paper.

% \begin{figure}[!htb]
% \centering
% \includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-4.png}
% \includegraphics[width=0.32\linewidth]{time-vs-ips-OS-4-24.png}
% \includegraphics[width=0.32\linewidth]{time-vs-ips-OS-7-31.png}
% \caption{Duration of experiments vs. number of different IPs (nodes)
%   participating in it, with averages and standard deviation shown as
%   black dots and lines, single dot if it was 1). Shade of gray or
%   color indicate how many experiments included that many unique IPs.
% From left to right, experiments 4/4, 4/24 and 7/31.}
% \label{fig:duration}
% \end{figure}
% % ../data/time-vs-IPs-openshift.R
%
\begin{figure}[!htb]
\centering
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-4.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-4-24.png}
\includegraphics[width=0.32\linewidth]{time-vs-rank-OS-7-31.png}
\caption{Duration of experiments vs. rank, with $y$ axis in a
  logarithmic scale. Dot color (or gray value) is related to the number of IPs
  participating in the experiment. From left to right, experiments
  4/4, 4/24 and 7/31.} 
\label{fig:zipf:os}
% computed via ../data/time-vs-IPs-openshift.R
\end{figure}
%
% We will have to analyze experimental data a bit further to find out why
% this happens and also if there are some patterns in the three sets of
% experiments. An interesting question to ask, for instance, is if
% by adding more computers makes the experiment take less. In fact, as
% shown in Figure \ref{fig:duration}, the {\em addition} of more computers does
% not seem to contribute to decreasing the time needed to finish the
% experiment. However, the cause-effect relationship is not clear at
% all. It might be the opposite: since experiments take longer to finish
% and might in fact be abandoned with no one contributing for some time,
% that increases the probability of someone new joining them. In fact,
% with experiments taking a few seconds and due to the way the
% experiments are announced, it is quite difficult that several
% volunteers join in in such a short period of time, even more if we take
% into account that volunteers are not {\em carried over} from previous
% experiments. This implies that it would be convenient to use a problem
% of a bigger size to check this hypothesis as we have done in the next
% experiment. 
%  However, at this point we
% have not found this convenient since there are several other issues
% that have to be solved, as it will be shown next. %Creo que este párrafo
%saldría sobrando ahora que hicmios un experimento más grande

@@

Although we think that, in general, the results presented in this
paper are independent of the problem chosen, it is true that even if
it is a difficult problem for evolutionary algorithms, it takes a few
seconds with the right settings to solve. This makes it difficult also
for this kind of asynchronous framework, but it would be interesting
to check the performance for other computationally demanding
tasks, especially ones that cover the middle ground between a problem
finishing in seconds or minutes (Trap) and weeks (Rastrigin F15). 
Besides, the two versions of the framework include as an
algorithmic variant using random population size. We do not think this
has had a big influence on the results and in fact this is not noticed
for F15, which is the second version tested, but it would be interesting to
measure exactly what this influence has been. 
% In general there are
% many issues with the evolutionary algorithm implementation itself,
% including using different, or adaptive, policies for inserting and
% sending individuals to the pool as was done in \cite{araujo2008mam},
% using different policies for population initialization, and also the
% incorporation of high-speed local resources to the pool to check what
% would be the real influence of the volunteer pool to the final
% performance.
